---
title: "Compare to other tools"
author: "Zuguang Gu ( z.gu@dkfz.de )"
date: '`r Sys.Date()`'
output: 
  html_document:
    css: main.css
    toc: true
---

```{r, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 6, fig.height = 6,
    fig.align = "center")
```

In this document, we compare **simona** to the following three R packages:
**ontologyIndex**/**ontologySimilarity**, **GOSemSim** and **GOSim**. Because
**GOSemSim** and **GOSim** only support Gene Ontology (GO), for this
comparison we take GO (the Biological Process namespace) as the test ontology.
**ontologyIndex** and **ontologySimilarity** belong to a suite of packages
which are called **ontologyX** where **ontologyIndex** provides data
structures and **ontologySimilarity** provides methods for semantic similarity
analysis.

Different tools may have different GO versions and different default
parameters. In the next section, we describe the differences of the four tools
and the parameters we selected to make the differences of parameters between
tools as small as possible.

**ontologyIndex**/**ontologySimilarity**, **GOSemSim** and **GOSim** mainly
support information contents (IC) calculated from gene annotation, so in the
comparison we will apply later, we only look at that IC method based on gene
annotations, taking human as the organism.

There might be functions having similar names in different packages. To reduce
the ambiguity, we add the `package::` prefix when calling functions in
corresponding packages.

## Description of tools

### simona

**simona** uses the **GO.db** package as the GO source and the
**org.Hs.eg.db** package as the gene annotation source. We take `"is_a"`,
`"part_of"` and `"regulates"` as the GO relation types. Note that because
`"regulates"` is a parent relation of `"negatively regulates"` and
`"positively regulates"`, the latter two are automatically included.

In the next code, we calculate the information content based on the gene
annotation on each GO term. On each GO term, **simona** takes the number of
**unique** genes annotated to the term and to all its offspring terms. If a GO
term has no gene annotated, its IC value is `NA`.

```{r}
library(simona)
dag = simona::create_ontology_DAG_from_GO_db(namespace = "BP", org_db = "org.Hs.eg.db", 
	relations = c("part_of", "regulates"))
ic1 = simona::term_IC(dag, method = "IC_annotation")

go_id1 = names(ic1[!is.na(ic1)])
```


### ontologyIndex and ontologySimilarity

**ontologySimilarity** provides a pre-computed object of ICs for all GO terms
(`ontologySimilarity::GO_IC`), but the GO data is very old (generated on
2016). In this comparison, to be consistent to **GO.db**, we regenerate the
ontology object with the `go-basic.obo` file downloaded from GO website
(http://release.geneontology.org/2023-01-01/ontology/index.html) with the same
version as in **GO.db** (2023-01-01). The gene annotation is from the
**org.Hs.eg.db** package.

On each term, **ontologyIndex** takes the number of **unique**
genes annotated to the term and to all its offspring terms. See in the source code: https://github.com/cran/ontologyIndex/blob/1f83db5178e1b2501fa681e0b4ca4c117f4c25ef/R/sets.R#L94.

```{r}
library(ontologyIndex)
onto = ontologyIndex::get_ontology("go-basic.obo", 
	propagate_relationships = c("is_a", "part_of", "regulates", "negatively_regulates", "positively_regulates"))

library(org.Hs.eg.db)
tb = toTable(org.Hs.egGO)
tb = tb[tb$Ontology == "BP", ]
lt = split(tb$go_id, tb$gene_id)

ic2 = ontologyIndex::get_term_info_content(onto, lt)
go_id2 = names(ic2)
```

### GOSemSim

**GOSemSim** uses **GO.db** package as the GO source and **org.Hs.eg.db** as
the gene annotation source. All relation types of `"is_a"`, `"part_of"`,
`"regulates"`, `"negatively regulates"` and `positively regulates` are
included.

On each GO term, **GOSemSim** takes the number of **all** genes annotated to
the term and to all its offspring terms **without taking the unique genes**.
Also note, a gene can be annotated to a GO term duplicatedly for multiple
evidence codes (e.g. the gene with Entrez ID "994" is annotated to a GO term
"GO:0000086" twice with evidence code "IBA" and "TAS",
`org.Hs.egGO2EG[["GO:0000086"]]`). If a GO term has no gene annotated, its IC
value is `Inf`.


```{r}
library(GOSemSim)
data = GOSemSim::godata("org.Hs.eg.db", ont = "BP")
ic3 = data@IC[is.finite(data@IC)]
go_id3 = names(ic3)
```

### GOSim

The settings is the same as **GOSemSim**. `GOSim::calcICs()` calculates the
information contents and save them to the object `IC` in a local file
`"ICsBPhumanall.rda"`.

```{r}
library(GOSim)
GOSim::setOntology("BP")
GOSim::calcICs()
load("ICsBPhumanall.rda")
ic4 = IC[is.finite(IC)]
go_id4 = names(ic4)
```

### Compare supported GO terms

We compare the supported GO terms in the four tools. The UpSet plot
shows all the four tools share almost the same set of GO terms, except
**GOSemSim** loses a tiny amount of terms. This is expected because all the
tools are based on the same version of GO data.

```{r, fig.width = 7, fig.height = 4}
lt = list(simona = go_id1, ontologyIndex = go_id2, GOSemSim = go_id3, GOSim = go_id4)
library(ComplexHeatmap)
UpSet(make_comb_mat(lt), column_title = "Numbers of supported GO terms")
```

We take the common GO terms in the four lists for the later analyses.

```{r}
go_id_common = intersect(intersect(go_id1, go_id2), intersect(go_id3, go_id4))
length(go_id_common)
```

## Compare Information contents

We simply make scatterplot for every pair of tools.

```{r, fig.width = 10, fig.height = 10}
df = data.frame(
	simona        = ic1[go_id_common],
	ontologyIndex = ic2[go_id_common],
	GOSemSim      = ic3[go_id_common],
	GOSim         = ic4[go_id_common]
)
pairs(df, pch = ".", col = "#00000040", xlim = c(0, 12), ylim = c(0, 12),
	main = "pairwise comparions of ICs")
```

**simona** and **ontologyIndex** almost calculate identical IC values.
**GOSemSim** and **GOSim** almost calculate identical IC values. This is
expected because **simona**/**ontologyIndex** use the same method for
calculating ICs, and **GOSemSim**/**GOSim** use another same method.
**simona** and **GOSemSim**, although not identical, still show very strong
linear relations, but with a systematic shift.

The difference between **simona** and **GOSemSim** is mainly due to how they
calculate the number of genes annotated to each GO term.

Due to the nature of the DAG structure of GO, if a GO term is annotated to a
gene, all its ancestor terms are also associated with that gene. The
calculation of annotated genes is applied in a recursive way.

**Method used by simona/ontologyIndex**: Denote $G^*_x$ as the set of genes
_directly_ annotated to the GO term $x$, and $G_x$ as the set of genes
annotated to $x$ after merging from its child terms, $G_x$ is the union of all
genes annotated to its child terms and the genes directly annotated to $x$.

$$ G_x = \left( \bigcup_{z \in \mathrm{C}_x} G_z \right) \bigcup G^*_x $$

where $\mathrm{C}_x$ is the set of $x$'s child terms.

The information content is:

$$ \mathrm{IC}(x) = -\log \left( \frac{|G_x|}{|G_r|} \right) $$

where $G_r$ is the set of all genes annotated to all GO terms (or to the root
of the GO DAG).

$G_x$ can also be written as:

$$ G_x = \bigcup_{z \in \mathrm{D}^+_x} G_z^* $$

where $\mathrm{D}^+_x$ is the set of $x$'s offspring terms plus $x$ itself. In
this way, $G_x$ is the union of all genes annotated to $x$ and $x$'s offspring
terms.

**Method used by GOSemsim/GOSim**: Since genes are not uniquified, we cannot
use the set denotations here. Let's denote $N^*_x$ as the number of genes
_directly_ annotated to term $x$, and $N_x$ is the number of genes annotated
to $x$ after merging from its child terms. Note that $N^*_x \ne |G^*_x|$
unless all genes directly annotated to $x$ are unique. $N_x$ is calculated as:

$$ N_x = N^*_x + \sum_{z \in \mathrm{C}_x} N_z $$

It can also be written as:

$$ N_x = \sum_{z \in \mathrm{D}^+_x} N^*_z $$

The information content is

$$ \mathrm{IC}(x) = -\log \left( \frac{N_x}{N_r} \right) $$

where $N_r$ is the value for the root term.

As shown in the scatter plot, **GOSemsim/GOSim** systematically over-estimate
the IC values. The two types of methods are the same when each gene is only
uniquely annotated to a single GO term.

## Compare semantic similarities


We randomly take 500 GO terms and calculate the semantic similarities among
them. For all the tools, we use the "Lin" method.

```{r}
set.seed(123)
go_id = sample(go_id_common, 500)

m1 = simona::term_sim(dag, go_id, method = "Sim_Lin_1998")
m2 = ontologySimilarity::get_term_sim_mat(ontology = onto, information_content = ic2, 
	method = "lin", row_terms = go_id, col_terms = go_id)
m3 = GOSemSim::termSim(go_id, go_id, semData = data, method = "Lin")
m4 = GOSim::getTermSim(go_id, method = "Lin")
```

We visualize the four matrices as heatmaps. Since they are symmetric and
correspond to the same set of GO terms, we align the four heatmaps to let them
have the same set of row orders and column orders.

```{r, fig.width = 18, fig.height = 4.5}
library(circlize)
# set the same color mapping to the four heatmaps
col_fun = colorRamp2(c(0, 0.3, 0.6), c("blue", "white", "red"))

od = hclust(dist(m1))$order

Heatmap(m1, name = "similarity", col = col_fun,
		show_row_names = FALSE, show_column_names = FALSE,
		row_order = od, column_order = od,
		column_title = "simona, random 500 GO terms") +
Heatmap(m2, name = "similarity", col = col_fun,
		show_row_names = FALSE, show_column_names = FALSE,
		row_order = od, column_order = od,
		column_title = "ontologySimilarity, random 500 GO terms") +
Heatmap(m3, name = "similarity", col = col_fun, 
		show_row_names = FALSE, show_column_names = FALSE,
		row_order = od, column_order = od,
		column_title = "GOSemSim, random 500 GO terms") +
Heatmap(m4, name = "similarity", col = col_fun, 
		show_row_names = FALSE, show_column_names = FALSE,
		row_order = od, column_order = od,
		column_title = "GOSim, random 500 GO terms")
```

**simona** and **ontologySimilarity** generate identical results. **simona**
and **GOSemSim** generate very similar similarities, except **GOSemSim**
generate systematic higher similarities than **simona**, which is because of
the higher ICs by **GOSemSim**. **GOSim** can somehow also show the "diagonal
block patterns" as the other three heatmps, but some GO terms have zero
similarities to all other GO terms, which makes the heatmap fragmented. A
closer look at **GOSim**'s source code reveals that **GOSim** has a different
way to look for the most informative common ancestor (MICA) terms.

In Lin's similarity method, the similarity between term $a$ and $b$ is
calculated as the IC of their MICA term $c$ normalized by the average of the IC
of them two:

$$ \mathrm{Sim}(a, b) = \frac{2 * \mathrm{IC}(c)}{\mathrm{IC}(a) + \mathrm{IC}(b)} $$

The MICA term $c$ is defined as:

$$ c = \operatorname*{arg\,max}_{t \in \mathrm{A}_a^+ \cap \mathrm{A}_b^+} \mathrm{IC}(t)  $$

where e.g. $\mathrm{A}_a^+$ is the set of $a$'s ancestors including $a$ itself.

Different in **GOSim**, the ancestor set excludes the term $a$ (see in the source code: https://github.com/cran/GOSim/blob/master/R/general.R#L24):

$$ c = \operatorname*{arg\,max}_{t \in \mathrm{A}_a \cap \mathrm{A}_b} \mathrm{IC}(t)  $$

This causes the problem where if $a$'s parent is the root $r$ which always has
$\mathrm{IC}(r) = 0$, the similarity between $a$ and any of $a$'s offspring
terms is always 0.


We can also perform a pairwise comparison on semantic similarities between
tools. We can draw the same conclusions as from the heatmaps.

```{r, fig.width = 10, fig.height = 10}
df = data.frame(
	simona             = m1[lower.tri(m1)],
	ontologySimilarity = m2[lower.tri(m2)],
	GOSemSim           = m3[lower.tri(m3)],
	GOSim              = m4[lower.tri(m4)]
)
pairs(df, pch = ".", col = "#00000040", xlim = c(0, 1), ylim = c(0, 1),
	main = "pairwise comparison of semantic similarities")
```

## Runtime

The key step in Lin's method is to look for MICA of every two terms. It is
also the dominant and most time-consuming part in the computation. In the code
below, we benchmark the runtime of the four tools on looking for MICA on
different numbers of terms.

Because **GOSim** has a bad runtime performance, we only run
`GOSim::getTermSim()` on less than 1000 terms.

```{r, eval = !file.exists("compare_tools.RData")}
set.seed(123)
k = seq(200, 10000, by = 200)
t1 = t2 = t3 = t4 = rep(NA_real_, length(k))
for(i in seq_along(k)) {
	go_id = sample(go_id_common, k[i])
	t1[i] = system.time(simona::term_sim(dag, go_id, method = "Sim_Lin_1998"))[3]
	t2[i] = system.time(ontologySimilarity::get_term_sim_mat(ontology = onto, information_content = ic2, 
		method = "lin", row_terms = go_id, col_terms = go_id))[3]
	t3[i] = system.time(GOSemSim::termSim(go_id, go_id, semData= data, method = "Lin"))[3]
	if(k[i] < 1000) t4[i] = system.time(GOSim::getTermSim(go_id, method = "Lin"))[3]
}
```

```{r, echo = FALSE}
if(!file.exists("compare_tools.RData")) {
	save(k, t1, t2, t3, t4, file = "compare_tools.RData")
} else {
	load("compare_tools.RData")
}
```

The following plot shows **simona** and **ontologySimilarity** show far better
performance than **GOSemSim** and **GOSim**. **simona** has the best
performance among the four tools.

```{r, fig.width = 6, fig.height = 6}
plot(NULL, xlim = c(0, 9000), ylim = c(0, 350), 
	xlab = "Number of terms", ylab = "runtime (sec)",
	main = "Compare runtime of the four tools")
lines(k, t1, col = 2)
lines(k, t2, col = 3)
lines(k, t3, col = 4)
lines(k, t4, col = 7)

legend("top", lty = 1, col = c(2, 3, 4, 7), 
	legend = c("simona", "ontologySimilarity", "GOSemSim", "GOSim"))
```

## Session info

```{r}
sessionInfo()
```


<script src="jquery.sticky.js"></script>
<script>
$(document).ready(function(){
    $("#TOC").sticky({
        topSpacing: 0,
        zIndex:1000    
    })
    $("#TOC").on("sticky-start", function() {

        $("<p style='font-size:1.2em; padding-left:4px;'><a id='TOC-click'>Table of Content</a></p>").insertBefore($("#TOC ul:first-child"));
        $("#TOC-click").hover(function() {
            $(this).css("color", "#0033dd").css("cursor", "pointer");
            $("#TOC").children().first().next().show();
            $("#TOC").hover(function() {
                $(this).children().first().next().show();
            }, function() {
                $(this).children().first().next().hide();
                $("body").off("hover", "#TOC");
            })
        }, function() {
            $(this).css("color", "#0033dd");
        })
        $("#TOC").children().first().next().hide();

    })
    $("#TOC").on("sticky-end", function() {
        $("#TOC").children().first().remove();
        $("#TOC").children().first().show();
    })
});
</script>

