---
title: "Compare IC methods"
author: "Zuguang Gu ( z.gu@dkfz.de )"
date: '`r Sys.Date()`'
output: 
  html_document:
    css: main.css
    toc: true
---

```{r, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 6, fig.height = 6,
    fig.align = "center")
```

**simona** supports many methods for calculating information contents (IC). In
this document, we compare different IC methods using Gene Ontology as the test
ontology.

By default, we use the Biological Process (BP) namespace in GO. We only take
`"is_a"` and `"part_of"` as the relation types. The `org_db` argument is also
set to human for the `IC_annotation` method.

```{r}
library(simona)
dag = create_ontology_DAG_from_GO_db(org_db = "org.Hs.eg.db")
dag
```

All IC methods supported in **simona** are listed as follows. The full
description of these IC methods can be found in the vignettes of **simona**.

```{r}
all_term_IC_methods()
```

We go over all IC methods in **simona** and save the results in a list.

```{r}
lt = lapply(all_term_IC_methods(), function(method) {
    term_IC(dag, method)
})
names(lt) = all_term_IC_methods()
df = as.data.frame(lt)
```

All the vector elements in `lt` contain IC values for all terms in the same order.

We calculate the correlations between IC vectors from different methods and
make the correlation heatmap. As 1 - correlation is also a dissimilarity
measurement, we directly generate the hierarchical clustering based on `1 -
cor`.

```{r, fig.width = 7}
cor = cor(df, use = "pairwise.complete.obs", method = "spearman")
hc = hclust(as.dist(1 - cor))

library(ComplexHeatmap)
Heatmap(cor, name = "correlation", cluster_rows = hc, cluster_columns = hc,
    column_title = "IC correlation, GO BP")
```

The heatmap shows the IC methods can be put into three groups. The groups can 
be manually extracted by observing sub-trees from the dendrogram.

```{r}
group = c("IC_annotation" = 1, "IC_Seddiqui_2010" = 1, "IC_Zhang_2006" = 1,
          "IC_offspring" = 1, "IC_Seco_2004" = 1, "IC_universal" = 2,
          "IC_Wang_2007" = 2, "IC_Meng_2012" = 3, "IC_Sanchez_2011" = 3,
          "IC_height" = 3, "IC_Zhou_2008" = 3)
```

The three groups of IC methods are:

1. `IC_offspring`, `IC_annotation`, `IC_Zhang_2006`, `IC_Seco_2004`,
  `IC_Seddiqui_2010`: Methods based on offspring terms.
2. `IC_universal`, `IC_Wang_2007`:
3. `IC_height`, `IC_Zhou_2008`, `IC_Sanchez_2011`, `IC_Meng_2012`: Methods based on the height of terms.


Next we perform MDS (multidimension scaling) analysis on the `1-cor` distance matrix and
visualize the first two dimensions. 

```{r, fig.width = 7}
library(ggrepel)
library(ggplot2)
loc = cmdscale(as.dist(1 - cor))
loc = as.data.frame(loc)
colnames(loc) = c("x", "y")
loc$method = rownames(loc)

loc$group = group[rownames(loc)]

ggplot(loc, aes(x, y, label = method, col = factor(group))) + 
    geom_point() + 
    geom_text_repel(show.legend = FALSE) +
    labs(x = "Dimension 1", y = "Dimension 2") +
    ggtitle("MDS based on the correlation between ICs")
```


We directly compare different IC methods with pairwise scatter plots, colored by 
depths of terms.

```{r, fig.width = 12, fig.height = 12}
pairs(df[, hc$order], pch = ".", col = dag_depth(dag), gap = 0)
```



```{r}
sessionInfo()
```
